---
title: "Class10"
author: "Ayse"
date: "2/18/2022"
output:
  pdf_document: default
  html_document: default
---

##Genotype data from 1000 gemones
We need to determine frequence of different alleles in the MXL population

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
table(mxl$Genotype..forward.strand.)/nrow(mxl)
```

In the GBR population
```{r}
mxl <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
table(mxl$Genotype..forward.strand.)/nrow(mxl)
```

##Homework
Q13:determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes

sample size for A/A is 108, A/G is 233. Median expression level for A/A is 31.25, A/G is 25.06, and G/G is 20.07.
```{r}
tbl <- read.table("rs8067378_ENSG00000172057.6.txt")
summary(tbl)
gg <- median(tbl[tbl$geno == "G/G",]$exp)
aa <- median(tbl[tbl$geno == "A/A",]$exp)
ag <- median(tbl[tbl$geno == "A/G",]$exp)
gg
aa
ag
```
boxplot for each genotype
```{r}
library(ggplot2)
bx <- ggplot(tbl, aes(geno, exp, fill = geno))+
         geom_boxplot(notch = "TRUE")+
         geom_point(geno,exp, color = geno)
bx
```

Q14: what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?
Expression of ORMDL3 is higher for  the A/A geontype compared to the G/G genotype. The SNP may down-regulate expression of ORMDL3.