---
title: "Class14"
author: "Ayse"
date: '2022-02-24'
output:
  html_document: default
  pdf_document: default
---
```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```


    Q1. What column details the total number of people fully vaccinated?
column 9
   
    Q2. What column details the Zip code tabulation area?
column 2

    Q3. What is the earliest date in this dataset?
January 5 2021

    Q4. What is the latest date in this dataset?
February 22, 2022
```{r}
#overview
#install.packages("skimr")
library(skimr)
skimr::skim(vax)
```

    Q5. How many numeric columns are in this dataset?
9

    Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?
18174

    Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 
17%
```{r}
fv_na <- sum( is.na(vax$persons_fully_vaccinated) )
(fv_na/length(vax$persons_fully_vaccinated))*100
```
#working with dates

Q9. How many days have passed since the last update of the dataset? 0
Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?  413
```{r}
#install.packages("tidyverse")
library(lubridate)
vax$as_of_date <- ymd(vax$as_of_date)
today () - vax$asofdate[1]
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```
 
#working with zipcodes
```{r}
#install.packages("zipcodeR")
library(zipcodeR)
geocode_zip('92037')
zip_distance('92037','92109')
reverse_zipcode(c('92037', "92109") )
```

    Q11. How many distinct zip codes are listed for San Diego County? 6420
    Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?


```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)

sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)

uni <- length(unique(sd))
```

Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-02-22”? 
```{r}
avg <- mean(sd$percent_of_population_fully_vaccinated)
```


Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-02-22”?